---
title: "Week 7: Example for dealing with dates in data"
---

# Claimer:
# This case study is taken from
# https://github.com/MeganBeckett/lubridate-heatmaps_tutorial and
# https://github.com/MeganBeckett/lubridate-heatmaps_tutorial/blob/master/code/running_casestudy.Rmd
# has been modified as needed.

---

# Introduction

This case follows on from investigating the lubridate package within 
the tidyverse and putting some of the features into practice 
by looking at running activity data over time.

We are interested to investigate some ways of plotting running activities 
over time. The data is a CSV export from Garmin Connect (https://connect.garmin.com/).

# Load packages needed

```{r message=FALSE, warning=FALSE}
# I am going to use several packages from tidyverse
# library(readr)
# library(dplyr)
# library(lubridate)
# library(ggplot2)
library(tidyverse)
```

# Read in data

```{r}
#When I use read_csv() function, I run into a problem with column names involving spaces.
#unfortunately, i could not solve this problem.
activities <- read_csv("data/Activities.csv")
View(activities)
```

```{r}
#not sure what is going on. but used this.
#https://community.rstudio.com/t/handling-column-names-from-df-with-spaces/58082/3
activities <- as_tibble(activities, .name_repair = "universal")
View(activities)
```

# Initial view of data

```{r}
dim(activities)
str(activities)
```

## Tidy up data
## Extract variables of interest

Many of the columns are empty, so let's create a smaller subset of those we are interested in.
Select the activity type, date and distance.
Convert Activity.Type to factor to see if there is more than one level.
Subset again to only look at data since the beginning of 2015

```{r}
#Create subset of variables
running_15 <- activities %>% 
              select(Activity.Type, Date, Distance) %>%                
              #Convert activity type to a factor
              mutate(Activity.Type = as.factor(Activity.Type)) %>% 
              filter(Date >= "2015-01-01")
str(running_15)
```

See levels explicitly

```{r}
running_15 %>% 
   pull("Activity.Type") %>% 
   levels()
```

########################################
### In the lecture start from here   ###
########################################

```{r}
View(running_15)
```


# Create new variables

Extract year, month, week, wday, hour components first,
then create new variables based on this.
Convert wday into a factor.

```{r}
running_long <- running_15 %>% 
                  mutate(year = year(Date), 
                         month = month(Date, label = TRUE, abbr = FALSE), #abbr=TRUE gives three-letter abb.
                         week = week(Date), 
                         wday = wday(Date, label = TRUE, abbr = FALSE, week_start = 1),
#label=TRUE:TRUE will display the day of the week as an "ordered factor" of character strings, such as "Sunday." 
#day on which week starts following ISO conventions - 1 means Monday, 7 means Sunday (default). 
#When label = TRUE, this #will be the first level of the returned factor.                          
                         hour = hour(Date))
str(running_long)  
```  

  
```{r}
running_long <- running_15 %>% 
                  mutate(year = year(Date), 
                         month = month(Date, label = TRUE, abbr = FALSE), #abbr=TRUE gives three-letter abb.
                         week = week(Date), 
                         wday = wday(Date, label = TRUE, abbr = FALSE, week_start = 1),
                         hour = hour(Date)) %>% 
#week_start	
#day on which week starts following ISO conventions - 1 means Monday
#turn ordered factor into a regular factor for compatibility in the following codes.
                  mutate(wday = factor(wday,
                         levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
                         ordered = FALSE))

```


```{r}
str(running_long)
```

# Exploratory visualisation

# Heatmap

We will build heatmaps using the geom_tile() layer in R ggplot2.
Input data must be a "long format" where each row provides an observation. 
At least 3 variables are needed per observation:
x: position on the X axis
y: position on the Y axis
fill: the numeric value that will be mapped to the color


## Weekly pattern - most popular times to run during the days of the week

We would like to see the most popular times to run during the days of the week
so 
x: hour, 
y: day of the week, 
fill: number of runs


```{r}
# Create a complete grid to include all possible combinations of weekday and hour even if no data
hour <- c(0:23)  #x-axis
week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday","Saturday", "Sunday") #y-axis
grid <- expand.grid(week, hour) 

#dim(grid): rows: 7*24=168 by 2
```

```{r}
# Group by week day, then hour and summarise by counting number of activities per group
heatmap_runs <- running_long %>%
                 group_by(wday, hour) %>%
                  summarise(no_runs = n())
#dim(heatmap_runs):75  3
```


```{r}
# Join grid and heatmap_runs
# If you don't do this, the heatmap won't show the full week, and have blank tiles for missing values instead of rather showing 0 runs/activity
heatmap_runs_full <- heatmap_runs %>%
                      right_join(grid, 
                                 by = c("wday"="Var1", "hour"="Var2")) %>% 
                      arrange(wday,hour)  

# right_join: includes all rows in grid data frame, takes the corresponding
# no runs from heatmap_runs data frame.
# dim(heatmap_runs_full): 168 x 3 
```

```{r}
#Only 75 out 168 will be filled in.
#Replace NA with 0 so that ggplot2 can handle (?)
heatmap_runs_full[is.na(heatmap_runs_full)] <- 0

#did not work. any help?
#heatmap_runs_full <- heatmap_runs_full %>% 
#                    mutate(no_runs = replace_na(no_runs, 0))
```


```{r}
# Plot heatmap using ggplot
heatmap_runs_full %>% 
ggplot(aes(x=hour, y=wday, fill=no_runs)) +
      geom_tile() +
      scale_fill_gradient(name = "No. of runs", low = "white", high = "blue") +
      #scale_fill_gradientn: creates a n-colour gradient from low to high. 
      scale_x_continuous(breaks = seq(0,23,1)) + 
      ggtitle("Heatmap showing weekly activity pattern") +
      xlab("Hour") +
      ylab("Week day") +
      theme_bw()
```

```{r}
#https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
# Using ColorBrewer to generate palette
#install RcolorBrewer packages first
library(RColorBrewer)
display.brewer.all()
```

```{r}
#pick 100 hundred different colors, list from light to dark. n=9 is fixed. number of different colors.
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))(100)
```

```{r}
heatmap_runs_full %>% 
ggplot(aes(x=hour, y=wday, fill=no_runs)) +
       geom_tile() +
       scale_fill_gradientn(name = "No. of runs", colours=c("white", pal)) +
       #colours, colors	Vector of colours to use for n-colour gradient.
       scale_x_continuous(breaks = seq(0,23,1)) + 
       ggtitle("Heatmap showing weekly activity pattern") +
       xlab("Hour") +
       ylab("Week day") +
       theme_bw()
```

#Save the last plot

```{r}
# Save plot
ggsave("heatmap_weekly_counts.png", width = 10, height = 7)
```

